---
title: "Creating maps in R - basics"
author: "Heidi Rautiainen"
date: "2024-12-20, Updated: 2026-01-21" 
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# rm(list = ls(all = TRUE))
# install.packages("pacman")
# library(pacman)
# pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
# pacman::p_loaded()

```

```{r packages, echo=T, include=FALSE}
# setwd("~/.../create-maps")
### install packages
# install.load::install_load("dplyr", "raster", "sf", "tidyverse", 
#                             "maps", "terra", "ggspatial")
knitr::opts_chunk$set(warning = TRUE, # show warnings
                      message = TRUE, # show messages
                      error = TRUE, # do not interrupt generation in case of errors,
                      echo = TRUE)

# Load packages
pcks <- list("dplyr", 
             "raster",    # used for raster data
             "terra",     # used for raster data 
             "sf",        # used for handling sppatial data 
             "ggspatial", # spatial for ggplot2 
             "tidyverse")

sapply(pcks, require, char = TRUE) 



# input files ----------------
spatial_folder <- "input_data"

# data to read
input_vector <- list.files(spatial_folder,
                          pattern = ".gpkg|.shp|.rds",
                          full.names = T,
                          recursive = T)

# data to read
maps_files_read <- list.files(spatial_folder,
                          pattern = ".tif",
                          full.names = T,
                          recursive = T)


```

# Exercise 1: Create study area map using "ggplot"

All datasets are found in the "data"-folder.

Here, we use the package 'rnaturalearth' to download countries in a ESRI shapefile format. These are available for download here: <https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-admin-0-countries/>

**Coordinate reference systems (CRS)**\
CRS provide a standardized way of describing locations to describe geographic data. The CRS that is chosen depends on when the data was collected, the purpose of the data, etc. It is necessary to transform vector and raster data to a common CRS when data with different CRS are combined.

***Setting CRS to SWEREF99 (Sweden). CRS can be referenced by its:***

1.  EPSG code, CRS("+init=epsg:3006") (see <http://www.epsg-registry.org/> and <http://spatialreference.org/>) or by

2.  proj4spring; "+proj=utm +zone=33 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

***There are two general options:***

1.  unprojected (a.k.a. Geographic): Latitude/Longitude for referencing location on the ellipsoid Earth (wgs84), and

2.  projected: Easting/Northing for referencing location on 2D representations of Earth (the creation of maps) e.g., SWEREF99.

More reading:\
<https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf>

## Load vector data

How to download using "rnaturalearth" and save the shp locally

```{r echo=F, include=FALSE}

# Download using 'rnaturalearth' package, or download from git -----------------
# install additional packages for high resolution data
# install.packages("rnaturalearthhires", repos = "https://ropensci.r-universe.dev")
# 
# library(rnaturalearth)
# 
# # # # shp of sweden (available in input_data/vector/country/swe.shp) -----
# swe_no <- ne_countries(scale = 10, country = c("Sweden", "Norway"))
# 
# # Check projection 
# sf::st_crs(swe_no)$proj4string #  unprojected 
# 
# class(swe_no)
# plot(swe_no)
# plot(swe_no[,1])
# 
# #transform projection to SWEREF
# swe_no_sweref <- sf::st_transform(swe_no, crs = CRS("+init=epsg:3006")) 
# Transform or convert coordinates of simple feature (sf)
# # can be set simply by "crs=3006". project(x, crs()) for raster
# 
# plot(swe_no_sweref[,1])
# class(swe_no_sweref)
# str(swe_no_sweref)
# st_crs(swe_no_sweref)$proj4string
# 
# # # Save as ESRI shp and load data:
# st_write(swe_no_sweref, "input_data/vector/country/swe_no.shp")
```

I will open the saved .shp for the exercise:

```{r echo=T}

## load country boarders -----------


#' downladed and saved from rnaturalearth (above)
swe_no_sweref <-  input_vector %>% 

  # value = T: return vector containing the matching elements
  grep(pattern = "swe_no", value = T) %>% 

  # read ESRI Shapefile object 
  sf::st_read() %>% 

  #convert geometry object into an sf object
  st_as_sf() 


# check coordinate system 
st_crs(swe_no_sweref)$proj4string

# Create a bounding box as an sf object for layers ----
b_box <- c(xmin = 350432, ymin = 7179284, 
           xmax = 857780, ymax =7444230)

# Now intersect the full world's coastline with the bounding box 
suppressWarnings({ 
sweden.c <- st_crop(swe_no_sweref, b_box)
plot(sweden.c)
}) 

class(sweden.c)
### water data ----

# import lakes
lakes <-  input_vector %>%
  grep(pattern = "lakes", value = T) %>%
  sf::st_read() %>%
  st_as_sf() 


class(lakes)
st_crs(lakes)

# Transform or convert coordinates of simple feature (sf)
lakes <- st_transform(lakes, crs = 3006) 
suppressWarnings({ 
  
# crop an sf object to a specific rectangle (here based on extent of "sweden.c"-shp)
# sf_use_s2(FALSE) # use if errors occur below
lakes <- st_crop(lakes, sweden.c) 

class(lakes)
plot(lakes)
}) 


```

### Add points of interest (cities)

Create xy points with labels. Here creating sf-object and setting the crs.

```{r echo=T}

# Create points of interest (coordinates from QGIS or google maps for demonstration)
places <- data.frame(ID= c("Arjeplog", "Arvidsjaur", "Sorsele"),
                     y = c(7328407.945, 7280856.153, 7270326.684),
                     x = c(630040.101, 693262.414, 617477.638)) %>%
  st_as_sf(coords = c("x", "y"),  # create sf object 
           crs=st_crs(sweden.c))  # set same crs as for sweden.c     
class(places)
# st_crs(places) # check projection 

```

## Make plot using ggplot

```{r}

library(ggspatial)

ggplot() +
  
    # plot sweden and lakes 
    geom_sf(aes(geometry = sweden.c$geometry)) + # geom_sf() for vector data 
  
    geom_sf(data = sweden.c,                     # plot sweden border vector
          aes(geometry = sweden.c$geometry),
          fill= "antiquewhite")+

      geom_sf(data = lakes,                      # plot lakes vector
              fill="lightblue") +

  # layer_spatial(data=elevation) +
  
  # add cities 
  geom_sf(data = places) +
  
  # add city names 
  geom_sf_text(data = places, aes(label = ID),
               vjust=-0.5)+
  
  # Add text on map
  annotate(geom = "text", 
             x = 484609, y = 7404230, 
             label = "Norway", 
             fontface = "italic", 
             color = "darkgray", 
           size = 6) +
  
    annotate(geom = "text", 
             x = 784609, y = 7404230, 
             label = "Sweden", 
             fontface = "italic", 
             color = "darkgray", 
           size = 6) +
  
  # Add north arrow
  annotation_north_arrow(location = "bl", 
                       which_north = "true", 
                         height = unit(1, "cm"),
  width = unit(1, "cm"),
                       pad_x = unit(0.5, "cm"), # horizontal align
                       pad_y = unit(8, "cm"), # vertical align
                       style = north_arrow_fancy_orienteering) +
  #add annotation scale
  annotation_scale(location = "bl", 
                     width_hint = 0.5) +
  
  
  # keep new projection for ggplot
  coord_sf(crs = st_crs(3006), 
  datum = sf::st_crs(3006),             
  expand = FALSE) +
  
  # add grid lines 
  theme(panel.grid.major = element_line(color = gray(.5), 
                                        linetype = "dashed", 
                                        linewidth  = 0.1), 
        panel.background = element_rect(fill = "lightblue")) 
    
### examplefor removing or adding titles 
# theme(axis.title.x = element_blank(),    # remove 
#       axis.title.y = element_blank()) + 
  # xlab("Longitude") + ylab("Latitude")   # add

ggsave("output_data/study_area.png", 
       width = 8, 
       height = 6, 
       dpi = 400)
```

Optional: Create a polygon and add to map

```{r ignore, echo=F}
#############################################################################
########      Optional: Create a polygon and add to map      ################
#############################################################################
# Create polygon for study area with following extent and add to the map ---
# 
#    xmin    ymin    xmax    ymax 
#  603379 7260836  740500 7352059 



# ######### Solution: 
# 
# df.box <- data.frame(x = c(min = 715694, max = 740824), 
#                      y = c(max = 7310002, max =7324832))
# 
# AOI  <- df.box %>% 
#   st_as_sf(coords = c("x", "y"), 
#            crs = st_crs(sweden.c)) %>% 
#   st_bbox() %>% 
#   st_as_sfc()
# 
# 
#      # add AOI to ggplot: 
#      geom_sf(data = AOI,
#              fill= "transparent",
#              linetype = "11",
#              color="red") +
```

\pagebreak

# Exercise 2: Maps in R

Pre-define crs using ESPG-code or proj4string by:

-   ESPG: CRS("+init=epsg:3006")

-   proj4string: "+proj=utm +zone=33 +ellps=GRS80 +units=m +no_defs"

```{r example maps}

# projection to use as proj4string 
crs_to_use <- "+proj=utm +zone=33 +ellps=GRS80 +units=m +no_defs" 

```

## Load background data

Now let's start by loading the maps using the "raster"-package.

Another option: "terra"-package using rast()-function)

```{r echo=FALSE}

# raster data ----
maps_files_read


maps <- raster::stack(maps_files_read)
# maps <- crop(maps, extent(AOI)) #when cropping is needed (not needed here)


names(maps)


names(maps) <- c("clear cuts", "elevation","distance to houses",
                 "lichen abundance", "land cover class", 
                 "distance to roads","slope", "terrain ruggedness index")

elevation <- maps[[2]]
sf::st_crs(elevation)$proj4string

# Plot elevation
plot(elevation,  main="Elevation")

# vector for home range ------- 
HR_area <- input_vector %>% 
  grep(pattern = "spatial_files_availability_PV_winter_2021_shorter_modified_new.shp", 
       value = T) %>% 
  
  sf::st_read() %>% 
  
  sf::st_transform(crs = crs_to_use) # set CRS

sf::st_crs(HR_area)$proj4string
plot(HR_area[,1], 
     main="Home range")

par(mfrow = c(1,1))

```

## Add animal data (GPS points)

Load prepared reindeer data

```{r echo=F}

reindeer_data <- readRDS("input_data/vector/animal_data/reindeer_data.rds") %>%
   dplyr::select(Collar_ID,  id, year, x_, y_, t_, group) 

str(reindeer_data)
# character to factor needed for grouping colors
reindeer_data$group <- as.factor(reindeer_data$group) 


# group individuals 
fed_ids <- reindeer_data %>% filter(group=="Fed")
ctrls <- reindeer_data %>% filter(group=="Control")


# create spatial points 
move <- SpatialPoints(reindeer_data[, c("x_", "y_")], 
                      proj4string = CRS("+proj=utm +zone=33 +ellps=WGS84 
                                        +datum=WGS84 +units=m +no_defs"))
class(move)
move_fed <- SpatialPoints(fed_ids[, c("x_", "y_")], 
                      proj4string = CRS("+proj=utm +zone=33 +ellps=WGS84 
                                        +datum=WGS84 +units=m +no_defs"))

move_ctrl <- SpatialPoints(ctrls[, c("x_", "y_")], 
                      proj4string = CRS("+proj=utm +zone=33 +ellps=WGS84 
                                        +datum=WGS84 +units=m +no_defs"))

# coordinates(move)
# proj4string(move)

library(mapview)

  move %>%
  mapview()
  
  
### subset for one ID
id<- reindeer_data %>% filter(Collar_ID=="FIT_421952")
id_points <- SpatialPoints(id[, c("x_", "y_")], proj4string = CRS("+proj=utm +zone=33 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"))

plot(id_points, 
          main="Animal position data (one ID)")
```

# Explore and select colors

```{r echo=F}

# Overview map terrain colour
over.col <- colorRampPalette(c("white", "black"))

plot(elevation)
par(mfrow = c(1,2))

plot(elevation, col=over.col(100),          
     main="Elevation: inverted")



plot(elevation, 
     col = gray.colors(100, 
                       start = 0.3, 
                       end = 0.9, 
                       gamma = 2.2, 
                       alpha = NULL),
          main="Elevation:  grey scale colors")


library(RColorBrewer)
par(mfrow = c(1,1))


display.brewer.all()
cols <- brewer.pal(7, "Greys")
pal <- colorRampPalette(cols[1:3])

head(terrain.colors(255))



plot(elevation,  
     main="Elevation")
```

## Prepare hillshade background map

Calculate slope and aspect from elevation (dem)

```{r echo=F}

par(mfrow = c(2,2))

# Create Hill shade
slope <- terra::terrain(elevation, 
                        "slope", 
                        unit='radians')

plot(slope,
          main="Slope")



aspect <- terra::terrain(elevation, 
                         "aspect", 
                         unit='radians')

plot(aspect, 
     main="Aspect")


hillshade <- hillShade(slope, #shade() for spatRaster
              aspect, 
              angle=45, 
              direction=300,   
              normalize = TRUE)

plot(hillshade,
               main="Hillshade")



# Plot hill shade and overlay DEM
plot(hillshade, col=rev(over.col(50)), legend=FALSE, 
                    main="Hillshade and overlay DEM")
plot(elevation, 
     col=adjustcolor(terrain.colors(100), 
                             alpha=0.75),      
     legend=FALSE, 
     add=TRUE)
```

## Plot and save all together

```{r echo=F}
# png("output/GPS_data.png", width = 30, height = 19, units = "cm", res = 400)
par(mfrow=c(2,2), mar=c(0.1, 0.1, 3, 0.1))


plot(hillshade, 
     col=rev(over.col(50)), 
          axes=FALSE,
         legend=FALSE)
plot(elevation, col=adjustcolor(terrain.colors(100), 
                             alpha=0.5), legend=FALSE, add=T)
     
     
title(main="A Plot fed inviduals",
      line = 0.5,
      cex.main=1.6,     
      adj = 0)
leg.txt <- c("Feeding group", "Control group")

points(move_fed, pch = 19, cex = 0.5,
       col = fed_ids$group)


# legend("topright", leg.txt, 
#        # inset = .01,
#        pch = 16, col = rev(1:2),
#        cex = 1.5, trace = TRUE)#
# dev.off()


#availbility 
plot(hillshade, 
     col=rev(over.col(50)), 
          axes=FALSE,
     legend=FALSE)
plot(elevation, col=adjustcolor(terrain.colors(100), 
                             alpha=0.5), legend=FALSE, add=T)
     

title(main="B Plot control individuals",
      line = 0.5,
      cex.main=1.6,     
      
      adj = 0)
points(move_ctrl, pch = 19, cex = 0.5, 
       col = ctrls$group)


#random 
plot(hillshade, 
     col=rev(over.col(50)), 
          axes=FALSE,
     legend=FALSE)
plot(elevation, col=adjustcolor(terrain.colors(100), 
                             alpha=0.5), legend=FALSE,add=T)
     

title(main="C Plot all individuals",
      line = 0.5,
      cex.main=1.6,     
      adj = 0)

points(move, pch = 19, cex = 0.5, 
       col = reindeer_data$group)


# all 

plot(hillshade, 
     col=rev(over.col(50)), 
          axes=FALSE,
     legend=FALSE)
plot(elevation, col=adjustcolor(terrain.colors(100), 
                             alpha=0.5), legend=FALSE, add=T)
     

title(main="D Plot home range",
      line = 0.5,
      cex.main=1.6,     
      adj = 0)
points(move, pch = 19, cex = 0.5, 
       col = reindeer_data$group)

plot(HR_area,   
     col = NA,
     border = 1,add=T)

# dev.off()

```

\pagebreak

# Saving it as ggplot

Some more packages required..

Credits to: <https://dominicroye.github.io/blog/hillshade-effect/index.html>

```{r echo = FALSE}
##```{r echo = FALSE, eval = FALSE}
library(terra)
library(ggnewscale)
library(tidyterra)
library(ggblend)

## not lakes in study area
# lakes2 <-  input_vector %>%
#   grep(pattern = "lakes", value = T) %>%
#   sf::st_read()
# st_crs(lakes2)
#
# lakes2 <- st_transform(lakes2, crs = st_crs(move))
# sf_use_s2(FALSE)
# lakes2 <- st_crop(lakes2,
#                   xmin = 715694, ymin = 7310002,
#                   xmax = 740824, ymax = 7324832)


extent(elevation)

#convert the raster to points for plotting
hillshade.p <- rasterToPoints(hillshade)
elevation.p <- rasterToPoints(elevation)

#Make the points a dataframe for ggplot
hillshade.df <- data.frame(hillshade.p)
elevation.df <- data.frame(elevation.p)


#Make appropriate column headings
names(hillshade.df)[3] <- "hillshade"
names(elevation.df)[3] <- "dem"



class(move)
move.sf <- st_as_sf(move)
move_fed.sf <- st_as_sf(move_fed)
move_ctrl.sf <- st_as_sf(move_ctrl)



# map simple hillshade --------------------

ggplot() +
  geom_raster(
    data = hillshade.df,
    aes(x, y, fill = hillshade),
    show.legend = FALSE
  ) +
  scale_fill_distiller(palette = "Greys") +
  new_scale_fill() +
  geom_raster(
    data = elevation.df,
    aes(x, y, fill = dem),
    alpha = .7
  ) +
  scale_fill_hypso_tint_c(breaks = c(
    180, 250, 500, 1000,
    1500, 2000, 2500, 3000, 3500
  )) +

  # add animal data
  geom_sf(data = move_fed.sf,
          colour="#B53737",
          size = 1) +
  geom_sf(data = move_ctrl.sf,
          colour="black",  size = 1)+


  # ### no lakes in the area
  # geom_sf(
  #   data = lakes2,
  #   fill = "#c6dbef", colour = NA
  # ) +
  #

  guides(fill = guide_colorsteps(
    barwidth = 20,
    barheight = .5,
    title.position = "right"
  )) +
  labs(fill = "m") +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom") + 
  labs(title="Simple hillshade using ggplot2")+
  theme(plot.title = element_text(hjust = 0.5))

```

\pagebreak

# More examples

## Multidimensional hillshade

```{r echo = FALSE}

# Multidirectional shadows ----------
slope.sr <- rast(slope) # make it spatr
aspect.sr <- rast(aspect) # make it spatr


# pass multiple directions to shade()
hillmulti <- map(c(270, 15, 60, 330),
                 function(dir) {
  shade(slope.sr,
            aspect.sr,
            angle = 45,
            direction = dir,
            normalize = TRUE)
                   }
  )


# create a multidimensional raster and reduce it by summing up
hillmulti <- rast(hillmulti) |> sum()

par(mfrow = c(1,1))

# multidirectional
plot(hillmulti, 
     col = grey(1:100 / 100), 
     main="Multidirectional hillshade")

hillshade.r <- rast(hillshade) # to spatrast to keep same format 
plot(hillshade.r, 
     col = grey(1:100 / 100), 
          main="Simple hillshade")



# map multidirectional hillshade --------------------
hillmulti.df <- as.data.frame(hillmulti, xy = TRUE)

# map
plot_map <- ggplot() +
  geom_raster(
    data = hillmulti.df,
    aes(x, y, fill = sum),
    show.legend = FALSE
  ) +
  scale_fill_distiller(palette = "Greys") +
  new_scale_fill() +
  geom_raster(
    data = elevation.df,
    aes(x, y, fill = dem),
    alpha = .7
  ) +

  scale_fill_hypso_tint_c(breaks = c(
    180, 250, 500, 1000,
    1500, 2000, 2500, 3000, 3500)) +
  
  # add animal data
  geom_sf(data = move_fed.sf,
          colour="#B53737",
          size = 1) +
  geom_sf(data = move_ctrl.sf,
          colour="black",  size = 1)+


  guides(fill = guide_colorsteps(
    barwidth = 20,
    barheight = .5,
    title.position = "right"
  )) +
  labs(fill = "m") +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom")+
    labs(title="Multidirectional hillshade using ggplot2")+
    theme(plot.title = element_text(hjust = 0.5))


plot_map
```

## Example of creating multiplots ggplots and to save the maps

```{r echo = FALSE, eval=FALSE}

# Example using blending
m <- ggplot() +
  list(
    geom_raster(
      data = hillmulti.df,
      aes(x, y, fill = sum),
      show.legend = FALSE
    ),
    scale_fill_distiller(palette = "Greys"),
    new_scale_fill(),
    geom_raster(
      data = elevation.df,
      aes(x, y, fill = dem),
      alpha = .7
    ),
    scale_fill_hypso_tint_c(breaks = c(
    180, 250, 500, 1000,
    1500, 2000, 2500, 3000, 3500
    ))
  ) |> blend("multiply") +


  guides(fill = guide_colorsteps(
    barwidth = 20,
    barheight = .5,
    title.position = "right"
  )) +
  
      # add animal data
  geom_sf(data = move_fed.sf,
          colour="#B53737",
          size = 1) +
  geom_sf(data = move_ctrl.sf,
          colour="black",  size = 1)+
  
  
  labs(fill = "m") +
  coord_sf() +
  theme_void() +
  theme(legend.position = "bottom")#+
    # labs(title="Multidirectional hillshade blended")+
    # theme(plot.title = element_text(hjust = 0.5))+
    # theme(plot.title = element_text(size = 20))


ggsave("output/hillshade_blend2.png", m,
  width = 10,
  height = 8,
  unit = "in",
  device = png,
  type = "cairo",
  bg = "white"
)

plot(m)


library(ggpubr)

fig <-ggpubr::ggarrange(

# Figure A
gpubr::ggarrange(ggarrange(plot_map,
                      labels = c("A"),
                      # common.legend = TRUE, legend = "bottom",
                      legend = FALSE,
                      hjust = -1,
                      vjust = 1.4),

# Figure B
         ggarrange(plot_map,
                      labels = c("B"),
                      # common.legend = TRUE, legend = "bottom",
                      legend = FALSE,
                      hjust = -1,
                      vjust = 1.4),
                ncol = 2,
                align = "h",
                common.legend = TRUE),

# Figure C
  ggarrange(ggarrange(plot_map,
                      labels = c("C"),
                      # common.legend = TRUE, legend = "bottom",
                      legend = FALSE,
                      hjust = -1,
                      vjust = 1.4),

            # Figure D
            ggarrange(plot_map,
                      labels = c("D"),
                      # common.legend = TRUE, legend = "bottom",
                      legend = FALSE,
                      hjust = -1,
                      vjust = 1.4),
            ncol = 2,
            align = "h",
            common.legend = FALSE),

  ncol = 1, nrow = 2,
  common.legend = TRUE)


fig2 <- annotate_figure(fig,
                        fig.lab.pos = c("top.left"),
                        top = text_grob("Figure 1", color = "Black", face = "bold", size = 22))
                              # top = text_grob("Visualizing len", color = "red", face = "bold", size = 14),
                              # bottom = text_grob("Data source: \n ToothGrowth data set", color = "blue",
                              #                    hjust = 1, x = 1, face = "italic", size = 10),
                              # left = text_grob("text", color = "black", size = 14, rot = 90, vjust = 1),
                              # bottom = text_grob("text", color = "black", size = 14))


# ggsave("output/GPS_data_ggplot.png",
#        plot = fig2,
#        width = 8, height = 6,
#          unit = "in",
#   device = png,
#   type = "cairo",
#   bg = "white",
#   dpi = 600)
# 
# ggsave("output/GPS_data_ggplot_example_multiplot.png", width = 8, height = 6, dpi = 400)

# save.image(file = "Tutorial.RData")

```
